\section{Related Work}

NE-emptiness is the main decision problem studied in this paper, i.e., decide if a given multiplayer concurrent game has a Nash Equilibrium or not. 
We studied this problem for $\LEX(\LTL,\mp)$-games, i.e., graph-games in which agents are trying to maximise their lexicographic payoff whose first co-ordinate is an \LTL formula and whose second co-ordinate is the mean-payoff of its weights. We proved the problem is $2$\exptime-complete.

\head{Other objectives}
NE-emptiness has been studied for graph-games with other objectives, a sample of which are given in Table~\ref{tab:NE-emptiness}.
\begin{table}\label{tab:NE-emptiness}
\begin{center} 
\begin{tabular}{llll}
Objectives			& Complexity 			& Reference\\
\hline
$\LEX(\LTL,\mp)$		& $2$\exptime-complete		& This paper\\
\LTL 				& $2$\exptime-complete		& \cite{DBLP:journals/tocl/MogaveroMPV14}\\
mean-payoff (\mp)			& \np-complete			& \cite{DBLP:journals/corr/abs-1109-6220}\\
reachability			& \np-complete			& \cite{DBLP:journals/corr/BouyerBMU15}\\
lexicographic reachability	& \pspace-complete		& \cite{DBLP:journals/corr/BouyerBMU15}\\
B\"uchi				& \ptime-complete		& \cite{DBLP:journals/corr/BouyerBMU15}\\
lexicographic B\"uchi		& \ptime-hard, in \np		& \cite{DBLP:journals/corr/BouyerBMU15}\\
% counting ???		& ???				& \cite{DBLP:journals/corr/BouyerBMU15}
% parity			& ${\sf P}^\np_{||}$-complete	& \cite{DBLP:journals/corr/BouyerBMU15}\\
\end{tabular}
\caption{Complexity of NE-emptiness for multiplayer concurrent-games with objectives from various classes.}
\end{center}
\end{table}
$\LEX(\LTL,\mp)$ generalises both mean-payoff games (simply set every formula to be $\true$) and \LTL games (simply set all the weights of the arena to be $0$). That NE-emptiness is decidable in $2$\exptime for \LTL games follows from a more general result, i.e., that model checking \emph{strategy logic} on concurrent game structures (i.e., arenas without the weights) is decidable \cite{DBLP:journals/tocl/MogaveroMPV14}. To get the $2$\exptime upper bound one writes the existence of a NE in a fragment of strategy logic (i.e., nested-goal with alternation number $1$) whose model checking problem is decidable in $2$\exptime. We cannot apply strategy logic to $\LEX(\LTL,\mp)$ games since the latter have a quantitative component while strategy logic, being built on top of \LTL, can only express qualitative objectives. 
Instead, as discussed in Section~\ref{sec:decision procedures}, we generalised the proof in~\cite{DBLP:journals/corr/abs-1109-6220} of the fact that NE-emptiness is in \np\ for mean-payoff games. 

We remark that the hardness results for lexicographic reachability and lexicographic B\"uchi are for lexicographic orders on an arbitrary (but finite) number of co-ordinates.

\head{Turn-based}
Very simple concurrent games, such as matching pennies, do not have NE. However, there are many classes of turn-based graph-games that always have NE. For instance, turn-based mean-payoff games~\cite{Alpe91}, turn-based games with $\omega$-regular objectives (which include \LTL objectives)~\cite{CJM04}, turn-based quantitative-reachability games~\cite{Brihaye2013}. Actually, this phenonemon has a general explanation: under certain conditions the existence of (finite-memory) NE for a class of objectives follows from the (finite-memory) determinacy of two-player zero-sum games with those objectives~\cite{DBLP:journals/corr/0001P16,Brihaye2013}. Although the latter citations do not cover the case of turn-based $\LEX(\LTL,\mp)$-games, we use a similar strategy in Theorem~\ref{thm:TB} to prove that these too always have NE. Determinacy (and the complexity of determining the winner) of two-player zero-sum games has been studied for a variety of objectives, 
i.e., mean-payoff~\cite{EM79,ZwPa95}, $\omega$-regular (which include $\LTL$-objectives)~\cite{DBLP:conf/dagstuhl/2001automata}.
%In addition, there are various uniform explanations for when such zero-sum games have positional or finite-memory winning strategies~\cite{Demri?,AmRu14,halfpositional}.

\head{Randomised Strategies}
We have considered deterministic strategies. Randomised strategies are functions that map a history to a probability distribution over actions. 
NE-emptiness is undecidable for randomised strategies on games with reachability objectives for each player~\cite{DBLP:conf/fsttcs/BouyerMS14}. Since reachability objectives are a special case of $\LEX(\LTL,\mp)$ objectives, we have undecidability of NE-emptiness for randomised strategies on $\LEX(\LTL,\mp)$ games.

\todo{IMPORTANT RELATED WORK THAT NEEDS TO BE SUMMARISED: cite{DBLP:journals/corr/abs-1303-0789}; as well as work on secure-Equilibria}

% \head{Other representations of games}
% Nash's famous result states that every finite game in strategic form has a randomised NE~\cite{??}. Computing such a NE is complete for the complexity class PPAD \cite{DBLP:journals/siamcomp/DaskalakisGP09}.\todo{check}
% 
% \todo{Stochastic games...}



 
% \begin{theorem} \cite{DBLP:journals/corr/abs-1109-6220}
%  The following problem is NP-complete: given $\bar{x} \leq \bar{z} \in \mathbb{Q}^{|Ag|}$ and a mean-payoff game, decide if there is a
%  NE, say with payoff $\bar{y}$, that satisfies $\bar{x} \leq \bar{y} \leq \bar{z}$.
%  \end{theorem}



% 
% \head{``Pure Nash Equilibria in Concurrent Deterministic Games'', Bouyer et al. \cite{DBLP:journals/corr/BouyerBMU15}}
% 
% The paper then gives the exact complexity of NE-emptiness for many $\omega$-regular 
% objectives, and semi-quantitative versions of these (like counting or 
% lexicographic). E.g., 
% \begin{theorem}[\cite{DBLP:journals/corr/BouyerBMU15}]
% The complexity of NE-emptiness  for:
% \begin{enumerate}
%  \item reachability- and safety-games are $\np$-complete, 
%  \item Buchi-games are $\ptime$-complete,
%  \item parity-games is ${\sf P}^\np_{||}$-complete,
%  \item Muller-games is $\pspace$-complete,
%  \item lexicographic Buchi-games are in $\np$.
% \end{enumerate}
% \end{theorem}

% \head{  Synthesis from LTL Specifications with Mean-Payoff Objectives`` \cite{DBLP:journals/corr/abs-1210-3539}}
% Two player, synthesis of specifications written in LTL with mean-payoff weights. 
% Objective is ''formula must be true and weights must be maximised``. Reduced to parity games.
% 
% \head{ A four-person chess-like game without Nash equilibria in pure stationary strategies}
% https://arxiv.org/abs/1411.0349
% 
% This paper shows that NE need not be memoryless for games in which the payoff depends on the
% first cycle formed. The example in their paper can be expressed in terms of meanpayoff.
% 
% 
% 
% \head{``Extending finite memory determinacy: General techniques and an application
%                to energy parity games'' \cite{DBLP:journals/corr/0001P16}}
%              
%              
%            Shows how to extend determinacy of win/lose games to finite-state NE of multiplayer games.
%            Turn based.
%            
%            
%       \input{oldnotes}