\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}

\parindent 0pt

\input{../macros}
%opening
\title{Complexities of Automata, Games, Synthesis}
\author{Sasha}

\begin{document}

\maketitle

\section{Model Checking}

\LTL, \CTLS \pspace

\CTL, \ptime

\section{Satisfiability}

\LTL, \pspace

\CTLS, ??

\CTL, ??

\section{Solving games}
States $n$, Transitions $m$

\subsection{Perfect Information}

Reachability $O(m)$ 

Buchi $O(mn)$ 

Parity $O(n^{d-1}m)$, $O((\frac{2n}{d})^{\lfloor \frac{d}{2} \rfloor} m)$

\LTL $2^n 2^{2^{O(|\psi|)}}$


\subsection{Imperfect Information}

Reachability $\exptime$ \cite{CDHR06, Reif}, belief-space

Buchi $\exptime$ \cite{CDHR06}, belief-space

Parity ?? nondeterministic exptime ??

\LTL $2$\exptime, \cite{KV}

\section{Formulas to automata} 
\LTL to ABW, linear 

\LTL to NBW,UCW $2^{O(|\phi|)}$ 


\CTL, alternation-free $\mu$-calculus to AWT, linear (KVW00)

\CTLS to ART, exponentially many states, ??linearly?? many pairs \cite{BVW94}

\CTLS to HAT, exponential states, linear depth KVW00

$\mu$-Calculus to ART, linear (EJ)

$\mu$-C to APT, linear size, number of colors is alternation depth, KVW00

\section{Automata to Automata}
$n$ states, $l$ priorities, $r$ pairs

NBW to  DPW, $2^n$ states, $n$ priorities \cite{Piterm06}

NBWs to DRW, $2^n$ states, $O(n)$ pairs (Safra)

APT (or 2APT) to NPT, $2^{nl\log(nl)}$ states, $O(nl)$ priorities \cite{Muller95,Vardi98}

ART to NRT, $(nr)^{O(nr)}$ states and $O(nr)$ pairs \cite{Muller95}

ABT to NBT, ??Exp?? \cite{Muller95}

\section{Membership of Automata}

NPT, $\np \cap co-\np$, equivalent to solving parity games, so $O(n^{d-1}m)$.

ART, \np \cite{KV}

AWT, linear \cite{KV}

\section{Emptiness of Automata}
$n$ states, $m$ size transition, $d$ degree of trees, $p$ priorities, $l$ input letters

DPW = parity game, $O(n^{p-1}m)$

NRW costs $(np)^{O(p)}$ \cite{PnueliRosner89}

NRT $(lnp)^{O(p)}$ \cite{PnueliRosner89}

NPT $O(dm(n+m)^p)$, \cite{Loding14}

NPT $(|\Sigma| n^{O(d)})^{O(p)}$, \cite{BastienThesis}

\subsection{Alternating}

APT = satisfiability of $\mu$-calculus, \exptime, (Chapter 9 of Automata, Logic, Games)

APT,AWT,AHT, fixed degree $d$, $\exptime$-c (survey by Vardi, Esparza; MSS86, MS95, FL79, Sei90)

\subsubsection{1-letter}

These reduce to emptiness of 1-letter alternating word automata with same condition, i.e., 2 player games. 

1APW = parity game = \np intersect co-\np

1AWW = linear time

1AHW = linear time, space $O(a \log^2b)$ where $a$ is depth and $b$ is size (see survey)


\section{Synthesis Perfect-Information}

Alphabets: Input $I$, Output $O$, Hidden environment $E$.

\subsection{\LTL}

$2$\exptime Lower Bound \cite{Rosner92}

$2$\exptime Upper Bound \cite{PnueliRosner89}



\emph{Method 1 - Orna-Moshe?}
\en
\- Translate $\Phi$ to a UCW by dualising the Vardi Wolper. This UCW has 1EXP many states.
\- Translate the UCW to a UCT by running it on all paths. No blowup.
\- Test the UCT for emptiness. The total cost is 2EXP.
\ne

\emph{Method 2 - Pnueli-Rosner}
\en
\- Translate $\Phi$ to a UBW by Vardi Wolper. This UBW has 1EXP many states.
\- Translate the UBW to a DRW by Safra. This DRW has 2EXP states, 1EXP rabin-pairs.
\- Translate the DRW to a DRT by running it on all paths. No blowup.
\- Test the DRT for emptiness. The total cost is $(2EXP 1EXP)^{O(1EXP)} = 2EXP$.
\ne



\subsection{\CTLS}

\subsection{\CTL}

$\exptime$ Lower from satisfiability \cite{FL97}

$\exptime$ Upper \cite{KV97}

\section{Synthesis Imperfect-Information}

\subsection{\LTL}

$2$\exptime Lower Bound from Perfect-Information.

$2$\exptime Upper Bound \cite{KVV97}

Build ABT that accepts strategy $\sigma:I^+ \to O$ iff every play consistent with $\sigma$ satisfies $\Phi$.
\en
\item Translate $\Phi$ to a UBW by Vardi Wolper. This UBW has 1EXP many states.
\item Build UBT that runs on $t_\sigma$, stores in its memory the belief-state, and universally launches the UBW in all directions. Blowup by size of largest belief-state, 
$2^{|E|}$ or $|E|^B$.
\item Emptiness of ABT is exponential. Total cost $2^{2^{O(|\varphi| + |E|)}}$ or $2^{2^{O(|\varphi|)} + |E|^B}$.
\ne

\subsection{\CTLS}

\subsection{\CTL}

\section{Automata Problem = Game Problem}

\subsection{Emptiness}

?? Right??

?? It seems these are equivalence. i.e., games can be thought of as emptiness problems of automata ??

\begin{enumerate}
 \item 
Nondet word emptiness = solving one player game.

\item 
Alt word emptiness = solving two player blind game ($\exists$-player does not see the state of the automaton) 

\item 
Nondet tree emptiness = solving two player game.

\item 
Alt tree emptiness = solving two player game of imperfect information in which observation sets correspond to directions (i.e., $\exists$-player does not see automaton-state, only tree-direction).

\end{enumerate}

\subsection{Membership}

Membership of Alternating Tree Automata are two player games.



\bibliographystyle{plain}
\bibliography{../../forsyte-publications/trunk/rubin.bib,../Bibfiles/chapters.bib,../Bibfiles/diss.bib,../Bibfiles/local.bib,../Bibfiles/main.bib,../Bibfiles/thesis.bib}

\end{document}
