% This is the abstract.  Max 2000 characters.\\

% \note{Proposal summary (identical to the abstract from the online proposal submission forms, section 1). 

% The abstract (summary) should, at a glance, provide the reader with a clear understanding of the objectives of the research proposal and how they will be achieved. The abstract will be used as the short description of your research proposal in the evaluation process and in communications to contact in particular the potential remote referees and/or inform the Commission and/or the programme management committees and/or relevant national funding agencies (provided you give permission to do so where requested in the online proposal submission forms, section 1). It must therefore be short and precise and should not contain confidential information. 

% Please use plain typed text, avoiding formulae and other special characters. The abstract must be written in English. There is a limit of 2000 characters (spaces and line breaks included).}

We are witnessing an increasing availability of \textbf{mechanisms}
that offer some form of programmability.
%%
These include software, manufacturing devices, smart objects and smart
spaces, intelligent robots, business process management systems, and
many others.
%%
All these mechanisms are being currently revolutionized by means of
advances in Machine Learning (ML).  In particular central sensing
capabilities (vision, language undestanding) and actuation
capabilities (robot-arm movement, camera twisting) are envisioned to be handled
by next generation ML-components.
%%
However, irrespective of the technological nature of the components, the connections between them are still
organized, monitored, and coordinated through standard programming.

\project (pronounced ``wisemech'') aims at developing the \textbf{science} and the \textbf{tools} for a new
generation of mechanisms to emerge: machanisms that are able to
\textbf{program themselves} without human intervention, and
automatically tailor their behavior so as to
%%
achieve desired goals,  maintain themselves within safe boundaries in a
changing environment, and keep following rules,
regulations and conventions that evolve over time. 
%%
% %Since with ``\emph{with great power comes great responsibility}'',
% Unlike ML approaches which are typically black-box, 
% \project intends to
% exploite \textbf{Knowledge Representation} (KR) for realizing self-programming mechanisms that are
% \textbf{white-box}: specifications and automatically synthesized
% programs must be human comprehensible. In
% other words, possibly on top of suitably characterized black-box 
% ML-components, the behavior enacted by \project self-programming
% mechanisms intends to be fully \textbf{explainable in human terms by design}.
%%
Self-programming calls for \textbf{balancing power and safety}.
\project intends to exploite \textbf{Knowledge Representation} (KR)
for realizing self-programming mechanisms that are \textbf{white-box},
that is, that are guarded and limited by human guided
specifications expressed in human compresible way. In other words, \project intends to develop
self-programming mechanisms, possibly on top of black-box
ML-components with suitable safety guarantees, whose specifications
and enacted behavior is fully \textbf{explainable in human terms by
  design}.


Scientifically, \project aims at integrating key ideas
%1
from \textbf{KR} on how to represent the domain of interest and 
mechanisms sensing and acting capabilities in a high-level human comprehensible
fashion,
%2
from \textbf{Data-aware Processes} in \emph{Databases} on how to
build data-aware dynamic behaviors,
%3
from \textbf{Verification and Synthesis} in \emph{Formal
  Methods} which provide mathematically elegant foundations for
synthesis,
%4
and from \textbf{Automated Planning} in \emph{AI}, which provide
algorithmic insights to the search process required by synthesis.
%
In this way \project will \textbf{bring together and cross-fertilize
  four research areas} with overlapping interests but developed by
\textbf{different scientific communities}.

\project will ground its scientific results upon diverse
\textbf{driving applications}, including manufacturing systems
(Industry 4.0), smart spaces (IoT) and business process management
(BPM).

 


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "PartB1"
%%% TeX-PDF-mode: t
%%% End:
