% This is the abstract.  Max 2000 characters.\\

% \note{Proposal summary (identical to the abstract from the online proposal submission forms, section 1). 

% The abstract (summary) should, at a glance, provide the reader with a clear understanding of the objectives of the research proposal and how they will be achieved. The abstract will be used as the short description of your research proposal in the evaluation process and in communications to contact in particular the potential remote referees and/or inform the Commission and/or the programme management committees and/or relevant national funding agencies (provided you give permission to do so where requested in the online proposal submission forms, section 1). It must therefore be short and precise and should not contain confidential information. 

% Please use plain typed text, avoiding formulae and other special characters. The abstract must be written in English. There is a limit of 2000 characters (spaces and line breaks included).}

We are witnessing an increasing availability of \textbf{mechanisms} that offer
some form of programmability.
%%
These include
software, % (CS), 
manufacturing devices, % (Industry 4.0), 
smart objects and smart spaces, % (IoT), 
intelligent robots, % (Robotics),
%autonomous agents (MAS)
business process models, % (BPM),
component-based systems, % (SOC and open APIs), 
and many others.
%%
All these mechanisms have some built-in capabilities that give them a
dynamic behaviour, which can be organized, refined and repurposed through
programming.

\project aims at allowing such mechanisms to \textbf{program
  themselves}, without human intervention.
%%
Through this self-programming ability such mechanisms can tailor their
behavior so as to
%%
achieve desired goals,  maintain themselves within a safe boundary in a
changing environment, and keep following rules,
regulations and conventions that evolve over time. 
%%

Unlike some machine learning approaches, the self-programming
mechanisms of interest in \project are \textbf{white-box}:
specifications and automatically synthesized programs must be human
comprehensible. In other words, we aim at obtaining self-programming
mechanisms whose behavior is fully \textbf{explainable in human terms}
by design.

\project's scientific work will merge key ideas 
%1
from \textbf{Knowledge Representation} in AI on how to represent the
system, its data and its properties in a high-level human
comprehensible fashion;
%2
from \textbf{Reactive Synthesis} in Formal Methods on providing
mathematically elegant foundations for synthesis, though focusing on
computationally more tractable formalisms recently proposed in
\textbf{Reasoning about Actions} in AI;
%3
from \textbf{Service Composition} in Service-Oriented Computing and
Open APIs and more recently in AI on how to compose, customize and
orchestrate modules and behaviors;
%
and
%4
from \textbf{Planning} in AI to gain algorithmic insight to the
synthesis process.



\project grounds its scientific results on diverse real  \textbf{application domains}, including business processes (BPM), manufacturing devices (Industry 4.0), and smart objects and spaces (IoT).%, and components-based systems (SOC and Open APIs).

 


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "PartB1"
%%% TeX-PDF-mode: t
%%% End:
